Radio Crickets: Chirping Jets from Black Hole Binaries Entering their Gravitational Wave Inspiral

### 2015-12-02 Paper/paper.tex
"Letter" is now "paper".

In this Letter, we study the morphology of jets from binary BHs whose
dynamics is doiminated by GW emission.  Jets in binary BHs have been
studied in a number of previous works \citep{1980ApJ...238L.129S,
  1981ApJ...246L..65I, 1981ApJ...246L.141H, 1982ApJ...258L..63G,
  1982ApJ...253L...1G, 1982ApJ...262..478G}.  It has been shown that
the orbital motion of binary results in a precession of the jet,
resulting in a helical or ``wiggly'' jet morphology.  Such models of
jet precession have been used to understand the observed jet
morphology in several AGN.  For example, recently
\citet{2014MNRAS.445.1370K} fit a jet precession model to the observed
jet in the quasar S5 1928$+$738 using 18 yr of very long baseline
interferometric (VLBI) data.  This object was first studied using BH
binary models by \citet{1993ApJ...409..130R}, but
\citet{2014MNRAS.445.1370K} also included effects of BH spin.  Other
objects studied in this manner are PG 1302-102
\citep{2015MNRAS.454.1290K, 2015Natur.518...74G}, BL Lacertae
\citep{2003MNRAS.341..405S, 2013MNRAS.428..280C}, NGC 4151
\citep{2012ApJ...759..118B}, OJ 287 \citep{2012MNRAS.421.1861V}, S5
1803$+$784 \citep{2008A&A...483..125R}, 3C 345
\citep{2005A&A...431..831L}, 3C 210 \citep{2004MNRAS.349.1218C}, PKS
420-014 \citep{2001A&A...374..784B}, 3C 273
\citep{2000A&A...360...57R}, and Mrk 501 \citep{1999A&A...347...30V}.
\citet{1982ApJ...262..478G} also studied several AGN using their model
of jet precession.

The shortest binary separations inferred from these studies are of the
order of $10^{-2}$ pc.  At smaller separations of around $10^{-3}$ pc,
the orbital velocity of the BHs typically becomes relativistic
($v_\mathrm{orb}\sim 10^4$ km$/$s or $\beta\sim 0.01$).  It was
thought that at these small separations, the binary evolves in a
vacuum, because the inspiral time scale of the binary is likely
shorter than the viscous time scale of any circumbinary gas disk.  As
a result, gas fails to catch up with the BHs, making any accretion
activity impossible in this stage of the binary's evolution
\citep{2005ApJ...622L..93M, 2010PhRvD..81b4019S, 2010ApJ...714..404T}.
However recent viscous hydrodynamical simulations do not find this
decoupling between the BHs and the circumbinary disk.  In these
simulations, the binary's member BHs can continue to accrete gas from
the circumbinary accretion disk all the way until coalescence
\citep{2012PhRvL.109v1102F, 2013MNRAS.436.2997D, 2014PhRvD..89f4060G,
  2014PhRvD..90j4030G, 2015MNRAS.446L..36F, 2015ApJ...807..131S}.
This opens up the possibility of jet production by one of the binary
members during the binary's final phase of evolution before
coalescence, i.e., when the binary's dynamics is dominated by GW
emission.  In this Letter we study the morphology of such jets.

@@

Further parameters can be introduced in the dynamical model by
considering unequal-mass binaries, eccentric orbits, and
Post-Newtonian corrections, but we not develop this further in this
Letter.  We make the code for implementing the model presented in this
Letter and for fitting it to observed jets using MCMC analyses
available at \url{https://github.com/gkulkarni/JetMorphology} under
the MIT open source software license.  This code can be used to
explore parameter space of the model.

@@

The jet morphology acts as an electromagnetic counterpart for chirping
gravitational wave sources.  BH binaries are expected to be the
brightest GW sources in the universe.  Current ground based GW
detectors such as LIGO are sensitive to intermediate-mass BH binaries
at high redshift, which would be difficult to detect with our method.
However, future detectors such as eLISA are sensitive to GW emission
from $10^7$--$10^9$ M$_\odot$ BHs \citep{2013CQGra..30x4009S}.  Pulsar
Timing Arrays are sensitive to even higher binary masses.  Monitoring
of jet morphologies for the features outlined in this letter will put
a lower bound on the expected abundance of bright GW
sources. Moreover, combined GW and electromagnetic observations of
compact binary mergers should enable detailed studies of astrophysical
processes in the strong-field GR regime \citep{2005ApJ...629...15H}.

>>>>

In this paper, we study the morphology of jets from binary BHs whose
dynamics is dominated by GW emission.  Jets in binary BHs have been
studied in a number of previous works \citep{1980ApJ...238L.129S,
  1981ApJ...246L..65I, 1981ApJ...246L.141H, 1982ApJ...258L..63G,
  1982ApJ...253L...1G, 1982ApJ...262..478G}.  It has been shown that
the orbital motion of binary results in a precession of the jet,
resulting in a helical or ``wiggly'' jet morphology.  Such models of
jet precession have been used to understand the observed jet
morphology in several AGN.  For example, recently
\citet{2014MNRAS.445.1370K} fit a jet precession model to the observed
jet in the quasar S5 1928$+$738 using 18 yr of very long baseline
interferometric (VLBI) data.  This object was first studied using BH
binary models by \citet{1993ApJ...409..130R}, but
\citet{2014MNRAS.445.1370K} also included effects of BH spin.  Other
objects studied in this manner are PG 1302-102
\citep{2015MNRAS.454.1290K, 2015Natur.518...74G}, BL Lacertae
\citep{2003MNRAS.341..405S, 2013MNRAS.428..280C}, NGC 4151
\citep{2012ApJ...759..118B}, OJ 287 \citep{2012MNRAS.421.1861V}, S5
1803$+$784 \citep{2008A&A...483..125R}, 3C 345
\citep{2005A&A...431..831L}, 3C 210 \citep{2004MNRAS.349.1218C}, PKS
420-014 \citep{2001A&A...374..784B}, 3C 273
\citep{2000A&A...360...57R}, and Mrk 501 \citep{1999A&A...347...30V}.
\citet{1982ApJ...262..478G} also studied several AGN using their model
of jet precession.

The shortest binary separations inferred from these studies are of the
order of $10^{-2}$ pc.  At smaller separations of around $10^{-3}$ pc,
the orbital velocity of the BHs typically becomes relativistic
($v_\mathrm{orb}\sim 10^4$ km$/$s or $\beta\sim 0.01$).  It was
thought that at these small separations, the binary evolves in a
vacuum, because the inspiral time scale of the binary is likely
shorter than the viscous time scale of any circumbinary gas disk.  As
a result, gas fails to catch up with the BHs, making any accretion
activity impossible in this stage of the binary's evolution
\citep{2005ApJ...622L..93M, 2010PhRvD..81b4019S, 2010ApJ...714..404T}.
However recent viscous hydrodynamical simulations do not find this
decoupling between the BHs and the circumbinary disk.  In these
simulations, the binary's member BHs can continue to accrete gas from
the circumbinary accretion disk all the way until coalescence
\citep{2012PhRvL.109v1102F, 2013MNRAS.436.2997D, 2014PhRvD..89f4060G,
  2014PhRvD..90j4030G, 2015MNRAS.446L..36F, 2015ApJ...807..131S}.
This opens up the possibility of jet production by one of the binary
members during the binary's final phase of evolution before
coalescence, i.e., when the binary's dynamics is dominated by GW
emission.  In this paper we study the morphology of such jets.

@@

Further parameters can be introduced in the dynamical model by
considering unequal-mass binaries, eccentric orbits, and
Post-Newtonian corrections, but we not develop this further in this
paper.  We make the code for implementing the model presented in this
paper and for fitting it to observed jets using MCMC analyses
available at \url{https://github.com/gkulkarni/JetMorphology} under
the MIT open source software license.  This code can be used to
explore parameter space of the model.

@@

The jet morphology acts as an electromagnetic counterpart for chirping
gravitational wave sources.  BH binaries are expected to be the
brightest GW sources in the universe.  Current ground based GW
detectors such as LIGO are sensitive to intermediate-mass BH binaries
at high redshift, which would be difficult to detect with our method.
However, future detectors such as eLISA are sensitive to GW emission
from $10^7$--$10^9$ M$_\odot$ BHs \citep{2013CQGra..30x4009S}.  Pulsar
Timing Arrays are sensitive to even higher binary masses.  Monitoring
of jet morphologies for the features outlined in this paper will put
a lower bound on the expected abundance of bright GW
sources. Moreover, combined GW and electromagnetic observations of
compact binary mergers should enable detailed studies of astrophysical
processes in the strong-field GR regime \citep{2005ApJ...629...15H}.
